
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```



```{r libraries, echo = TRUE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(sf)
library(INLA)
library(geodata)
library(terra)
```


```{r load-data}

epsg = 3857 #4326
epsg_crs = "EPSG:3857"#Studt area - Quebec

# Study area
study_area0 <- st_read("C:/Users/maria/Documents/Chapter1/bayesian_model/data/study_area/canada.shp")
#study_area0 <- st_read("C:/Users/maria/Documents/Chapter1/bayesian_model/data/study_area/quebec.shp")
study_area <- study_area0 |>
  st_transform(epsg) |>
  st_union()|>
  st_simplify(dTolerance = 1000)|> # 1 km tolerance in EPSG:3857
  st_make_valid()

# Occurrence data
monarch_data <- st_read("C:/Users/maria/Documents/Chapter1/bayesian_model/data/species/PA_monarch_CAN.shp") #en epsg 3857
## For sampling

monarch_sf <- monarch_data |>
  st_transform(crs = epsg)

# Expert knowledge data
ek_pnt <- read.csv("C:/Users/maria/Documents/Chapter1/bayesian_model/data/expert_knowledge/Points.csv")
ek_answrs <- read.csv("C:/Users/maria/Documents/Chapter1/bayesian_model/data/expert_knowledge/Responses.csv")

# Environmental layers - Ouranos
tas_mean0 <- rast("C:/Users/maria/Documents/Chapter1/bayesian_model/data/predictors/tas_mean.tif")
pr_mean0 <- rast("C:/Users/maria/Documents/Chapter1/bayesian_model/data/predictors/pr_mean.tif")

```


```{r prepare-env-layers}
tas_mean <- project(tas_mean0, epsg_crs)
tas_mask <- mask(tas_mean, vect(study_area))
mt <- global(tas_mask, mean, na.rm = TRUE)[1,1]
st <- global(tas_mask, sd, na.rm = TRUE)[1,1]
std_tas <- (tas_mask - mt) / st


pr_mean <- project(pr_mean0, epsg_crs)
pr_mask <- mask(pr_mean, vect(study_area))
mp <- global(pr_mask, mean, na.rm = TRUE)[1,1]
sp <- global(pr_mask, sd, na.rm = TRUE)[1,1]
std_pr <- (pr_mask - mp) / sp


env <- c(std_tas, std_pr)
names(env) <- c("temp", "prec")

env

env_df <- as.data.frame(env, xy = TRUE, na.rm = FALSE)

# Track which cells are non-NA
valid_idx <- complete.cases(env_df[, c("temp", "prec")])
env_valid <- env_df[valid_idx, ]
```


```{r expert-knowledge}

# Spatialize expert points
sf_pnt <- ek_pnt |>
  st_as_sf(coords = c("long", "lat"),
           crs = 4326) |>
  st_intersection(study_area0) |>
  st_transform(crs = epsg) 
head(sf_pnt, 5)

sf_pnt <- sf_pnt |>
  mutate(y = as.integer(type == "presence"),
         temp = terra::extract(env$temp, vect(sf_pnt))[,2],
         prec = terra::extract(env$prec, vect(sf_pnt))[,2])
# standarize range values and Keep the necessary columns
answrs_select <- ek_answrs |>
  mutate(std_min_temp = (min.temp - mt) / st,
         std_max_temp = (max.temp - mt) / st,
         std_min_prec = (min.prec - mp) / sp,
         std_max_prec = (max.prec - mp) / sp) |>
  select("survey_id", "std_min_temp", "std_max_temp", "confidence.temp", "std_min_prec", "std_max_prec", "confidence.prec")

```

Using Stan to run step 1, expert informed model

```{r EK-stan-data}
#library(cmdstanr)
library(posterior)
library(loo)
library(rstan)

sigma_levels <- c(low = 0.01, medium = 0.5, high = 1.0)
sigma_levels

stan_data <- list(
  N  = length(sf_pnt$y),
  y  = as.integer(sf_pnt$y),
  x1 = sf_pnt$temp,
  x2 = sf_pnt$prec,

  # Covariate 1 expert values + uncertainty
  a1_star = answrs_select$std_min_temp,
  b1_star = answrs_select$std_max_temp,  
  c1_star = (answrs_select$std_max_temp/2),
  sigma_a1 = sigma_levels["medium"],
  sigma_b1 = sigma_levels["low"],
  sigma_c1 = sigma_levels["medium"],

  # Covariate 2 expert values + uncertainty
  a2_star = answrs_select$std_min_prec,  
  b2_star = answrs_select$std_max_prec,  
  c2_star = (answrs_select$std_max_prec/2),
  sigma_a2 = sigma_levels["low"],
  sigma_b2 = sigma_levels["low"],
  sigma_c2 = sigma_levels["low"],

  # Slope priors (both positive effects expected)
  beta1_mu = 1, 
  beta1_sigma = sigma_levels["medium"],
  beta2_mu = 1, 
  beta2_sigma = sigma_levels["medium"]
)


```



```{r EK-model-stan}

stan_code <- stan_model(file = "Step1-Expert_model.stan")


fit <- sampling(
  stan_code,
  data = stan_data,
  chains = 4, 
  refresh = 10000
)

```


```{r EK-predict-function}
predict_occurrence <- function(fit, new_x1, new_x2) {
  
  draws <- as.matrix(fit, pars = c("alpha", "beta1", "beta2",
                                    "a1", "c1", "b1",
                                    "a2", "c2", "b2"))
  
  tri <- function(x, a, b, c) {
    ifelse(x < a | x > b, 0,
      ifelse(x <= c,
        2 * (x - a) / ((b - a) * (c - a)),
        2 * (b - x) / ((b - a) * (b - c))))
  }
  
  preds <- sapply(seq_along(new_x1), function(i) {
    f1 <- tri(new_x1[i], draws[, "a1"], draws[, "b1"], draws[, "c1"])
    f2 <- tri(new_x2[i], draws[, "a2"], draws[, "b2"], draws[, "c2"])
    mu <- draws[, "alpha"] + draws[, "beta1"] * f1 + draws[, "beta2"] * f2
    mean(plogis(mu))
  })
  
  preds
}

preds <- predict_occurrence(fit,
                            new_x1 = env_valid$temp,
                            new_x2 = env_valid$prec)
```


```{r EK-model}
# Create an empty raster with same extent/resolution
pred_raster1 <- rast(env[[1]])  
values(pred_raster1) <- NA

# Fill in predictions at valid cells
pred_raster1[valid_idx] <- preds

non_na_extent1 <- ext(trim(pred_raster1))
pred_cropped1 <- crop(pred_raster1, non_na_extent1)

```


```{r EK-sd}
predict_sd <- function(fit, new_x1, new_x2) {
  draws <- as.matrix(fit, pars = c("alpha", "beta1", "beta2",
                                    "a1", "c1", "b1",
                                    "a2", "c2", "b2"))
  tri <- function(x, a, b, c) {
    ifelse(x < a | x > b, 0,
      ifelse(x <= c,
        2 * (x - a) / ((b - a) * (c - a)),
        2 * (b - x) / ((b - a) * (b - c))))
  }
  
  sds <- sapply(seq_along(new_x1), function(i) {
    f1 <- tri(new_x1[i], draws[, "a1"], draws[, "b1"], draws[, "c1"])
    f2 <- tri(new_x2[i], draws[, "a2"], draws[, "b2"], draws[, "c2"])
    mu <- draws[, "alpha"] + draws[, "beta1"] * f1 + draws[, "beta2"] * f2
    sd(plogis(mu))  # <-- sd instead of mean
  })
  sds
}

sd_raster1 <- rast(env[[1]])
values(sd_raster1) <- NA
sd_raster1[valid_idx] <- predict_sd(fit, env_valid$temp, env_valid$prec)

non_na_extent1 <- ext(trim(sd_raster1))
sd_cropped1 <- crop(sd_raster1, non_na_extent1)

```



```{r 70stan-summary}
# Extract mean and SD from Stan posterior to use as INLA priors
stan_fixed <- summary(fit, pars = c("alpha", "beta1", "beta2"),
                      probs = c(0.025, 0.5, 0.975))$summary

# Pull mean and sd for each parameter
prior_alpha_mean  <- stan_fixed["alpha",  "mean"]
prior_alpha_sd    <- stan_fixed["alpha",  "sd"]

prior_beta1_mean  <- stan_fixed["beta1",  "mean"]
prior_beta1_sd    <- stan_fixed["beta1",  "sd"]

prior_beta2_mean  <- stan_fixed["beta2",  "mean"]
prior_beta2_sd    <- stan_fixed["beta2",  "sd"]

```



```{r 70prior}

# Convert SD to precision
prec_alpha <- 1 / prior_alpha_sd^2
prec_beta1 <- 1 / prior_beta1_sd^2
prec_beta2 <- 1 / prior_beta2_sd^2

# Build INLA-format priors from Stan posteriors
prior_alpha <- list(prior = "gaussian",
                    param = c(prior_alpha_mean, prec_alpha))

prior_beta1 <- list(prior = "gaussian",
                    param = c(prior_beta1_mean, prec_beta1))

prior_beta2 <- list(prior = "gaussian",
                    param = c(prior_beta2_mean, prec_beta2))


pedge1 <- 0.01
edge1 <- min(c(diff(st_bbox(study_area)[c(1,3)])*pedge1,diff(st_bbox(study_area)[c(2,4)])*pedge1))


mesh <- inla.mesh.2d(loc.domain = NULL,
                     max.edge = c(edge1,edge1*3),
                     min.angle = 21,
                     cutoff = edge1/1,
                     offset = c(edge1,edge1*3),
                     boundary = study_area, 
                     crs = st_crs(study_area))


spde <- inla.spde2.pcmatern(
  mesh  = mesh,
  #prior.range = c(500000, 0.05),
  #prior.sigma = c(1, 0.01)
  prior.range = c(10, 0.05),   
  prior.sigma = c(1, 0.05)    
)
iset  <- inla.spde.make.index("s", n.spde = spde$n.spde)
formula <- y ~ -1 + intercept + x1 + x2 + f(s, model = spde)

```



# 70% simulation
## Update model with INLA - expert informed

```{r 70}
# ----
removal_pct   <- 70
removal_pct_txt <- gsub("\\.", "_", as.character(removal_pct))
monarch_sf <- st_read(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/data/species/simulation/PA/", "SF_", removal_pct_txt, "_monarch.shp"))
# ----
```


## INLA model - expert informed priors


```{r 70join-env-occ}

monarch_occ <- monarch_sf |>
  mutate(y = monarch_sf$presenc,
         temp = terra::extract(env$temp, vect(monarch_sf))[,2],
         prec = terra::extract(env$prec, vect(monarch_sf))[,2]) |>
  select(c("y","temp","prec"))

head(monarch_occ, 5)

```


```{r 70inla_df}

inla_df <- data.frame(
  y  = as.integer(monarch_occ$y),   
  x1 = monarch_occ$temp,       
  x2 = monarch_occ$prec
)


```


```{r 70spde}
coords <- as.matrix(sf::st_coordinates(monarch_occ))

A_est <- inla.spde.make.A(mesh, loc = coords)

stack_est <- inla.stack(
  data    = list(y = inla_df$y),
  A       = list(A_est, 1),
  effects = list(iset,
                 data.frame(intercept = 1,
                            x1 = inla_df$x1,
                            x2 = inla_df$x2)),
  tag     = "est"
)

coords_pred <- as.matrix(env_valid[, c("x", "y")])

# Projector matrix for prediction locations
A_pred <- inla.spde.make.A(mesh, loc = coords_pred)

stack_pred <- inla.stack(
  data    = list(y = NA),
  A       = list(A_pred, 1),
  effects = list(
    iset,
    data.frame(intercept = 1,
               x1 = env_valid$temp,
               x2 = env_valid$prec)
  ),
  tag = "pred"
)

# Join estimation + prediction stacks
stack_full <- inla.stack(stack_est, stack_pred)

```



```{r 70pred-INLA}

# Refit on full stack
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),

  control.fixed = list(
    mean = list(intercept = prior_alpha_mean,
                x1        = prior_beta1_mean,
                x2        = prior_beta2_mean),
    prec = list(intercept = prec_alpha,
                x1        = prec_beta1,
                x2        = prec_beta2)
  ),

  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]
```


```{r 70Plot-INLA}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Model update - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)

png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Model update - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()
```


# INLA model - non informed priors

```{r 70prior-NonInfo}

prior.beta <- list(prec = list(default = 1/(1)^2, Intercept = 1/(20)^2), 
                   mean = list(default = 0, Intercept = 0))

```



```{r 70full-NonInfo}
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),
  control.fixed = prior.beta,
  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]


```


```{r 70Plot--NonInfo}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Non informative - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Non informative - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()
```


# 90% simulation
```{r 90}
# ----
removal_pct   <- 90
removal_pct_txt <- gsub("\\.", "_", as.character(removal_pct))
monarch_sf <- st_read(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/data/species/simulation/PA/", "SF_", removal_pct_txt, "_monarch.shp"))
# ----
```


## INLA model - expert informed priors


```{r 90join-env-occ}

monarch_occ <- monarch_sf |>
  mutate(y = monarch_sf$presenc,
         temp = terra::extract(env$temp, vect(monarch_sf))[,2],
         prec = terra::extract(env$prec, vect(monarch_sf))[,2]) |>
  select(c("y","temp","prec"))

head(monarch_occ, 5)

```


```{r 90inla_df}

inla_df <- data.frame(
  y  = as.integer(monarch_occ$y),   
  x1 = monarch_occ$temp,       
  x2 = monarch_occ$prec
)


```


```{r 90spde}
coords <- as.matrix(sf::st_coordinates(monarch_occ))

A_est <- inla.spde.make.A(mesh, loc = coords)

stack_est <- inla.stack(
  data    = list(y = inla_df$y),
  A       = list(A_est, 1),
  effects = list(iset,
                 data.frame(intercept = 1,
                            x1 = inla_df$x1,
                            x2 = inla_df$x2)),
  tag     = "est"
)

coords_pred <- as.matrix(env_valid[, c("x", "y")])

# Projector matrix for prediction locations
A_pred <- inla.spde.make.A(mesh, loc = coords_pred)

stack_pred <- inla.stack(
  data    = list(y = NA),
  A       = list(A_pred, 1),
  effects = list(
    iset,
    data.frame(intercept = 1,
               x1 = env_valid$temp,
               x2 = env_valid$prec)
  ),
  tag = "pred"
)

# Join estimation + prediction stacks
stack_full <- inla.stack(stack_est, stack_pred)

```



```{r 90pred-INLA}

# Refit on full stack
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),

  control.fixed = list(
    mean = list(intercept = prior_alpha_mean,
                x1        = prior_beta1_mean,
                x2        = prior_beta2_mean),
    prec = list(intercept = prec_alpha,
                x1        = prec_beta1,
                x2        = prec_beta2)
  ),

  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]
```


```{r 90Plot-INLA}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Model update - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)

png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Model update - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()
```


# INLA model - non informed priors

```{r 90prior-NonInfo}

prior.beta <- list(prec = list(default = 1/(1)^2, Intercept = 1/(20)^2), 
                   mean = list(default = 0, Intercept = 0))

```



```{r 90full-NonInfo}
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),
  control.fixed = prior.beta,
  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]


```


```{r 90Plot--NonInfo}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Non informative - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Non informative - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()
```

# 99% simulation
## Update model with INLA - expert informed

```{r 99}
# ----
removal_pct   <- 99
removal_pct_txt <- gsub("\\.", "_", as.character(removal_pct))
monarch_sf <- st_read(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/data/species/simulation/PA/", "SF_", removal_pct_txt, "_monarch.shp"))
# ----
```


## INLA model - expert informed priors


```{r 99join-env-occ}

monarch_occ <- monarch_sf |>
  mutate(y = monarch_sf$presenc,
         temp = terra::extract(env$temp, vect(monarch_sf))[,2],
         prec = terra::extract(env$prec, vect(monarch_sf))[,2]) |>
  select(c("y","temp","prec"))

head(monarch_occ, 5)

```


```{r 99inla_df}

inla_df <- data.frame(
  y  = as.integer(monarch_occ$y),   
  x1 = monarch_occ$temp,       
  x2 = monarch_occ$prec
)


```


```{r 99spde}
coords <- as.matrix(sf::st_coordinates(monarch_occ))

A_est <- inla.spde.make.A(mesh, loc = coords)

stack_est <- inla.stack(
  data    = list(y = inla_df$y),
  A       = list(A_est, 1),
  effects = list(iset,
                 data.frame(intercept = 1,
                            x1 = inla_df$x1,
                            x2 = inla_df$x2)),
  tag     = "est"
)

coords_pred <- as.matrix(env_valid[, c("x", "y")])

# Projector matrix for prediction locations
A_pred <- inla.spde.make.A(mesh, loc = coords_pred)

stack_pred <- inla.stack(
  data    = list(y = NA),
  A       = list(A_pred, 1),
  effects = list(
    iset,
    data.frame(intercept = 1,
               x1 = env_valid$temp,
               x2 = env_valid$prec)
  ),
  tag = "pred"
)

# Join estimation + prediction stacks
stack_full <- inla.stack(stack_est, stack_pred)

```



```{r 99pred-INLA}

# Refit on full stack
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),

  control.fixed = list(
    mean = list(intercept = prior_alpha_mean,
                x1        = prior_beta1_mean,
                x2        = prior_beta2_mean),
    prec = list(intercept = prec_alpha,
                x1        = prec_beta1,
                x2        = prec_beta2)
  ),

  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]
```


```{r 99Plot-INLA}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Model update - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)

png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Model update - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()
```


# INLA model - non informed priors

```{r 99prior-NonInfo}

prior.beta <- list(prec = list(default = 1/(1)^2, Intercept = 1/(20)^2), 
                   mean = list(default = 0, Intercept = 0))

```



```{r 99full-NonInfo}
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),
  control.fixed = prior.beta,
  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]


```


```{r 99Plot--NonInfo}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Non informative - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Non informative - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()
```


# 99.9% simulation
## Update model with INLA - expert informed
```{r 999}
# ----
removal_pct   <- 99.9
removal_pct_txt <- gsub("\\.", "_", as.character(removal_pct))
monarch_sf <- st_read(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/data/species/simulation/PA/", "SF_", removal_pct_txt, "_monarch.shp"))
# ----
```


## INLA model - expert informed priors


```{r 999join-env-occ}

monarch_occ <- monarch_sf |>
  mutate(y = monarch_sf$presenc,
         temp = terra::extract(env$temp, vect(monarch_sf))[,2],
         prec = terra::extract(env$prec, vect(monarch_sf))[,2]) |>
  select(c("y","temp","prec"))

head(monarch_occ, 5)

```


```{r 999inla_df}

inla_df <- data.frame(
  y  = as.integer(monarch_occ$y),   
  x1 = monarch_occ$temp,       
  x2 = monarch_occ$prec
)


```


```{r 999spde}
coords <- as.matrix(sf::st_coordinates(monarch_occ))

A_est <- inla.spde.make.A(mesh, loc = coords)

stack_est <- inla.stack(
  data    = list(y = inla_df$y),
  A       = list(A_est, 1),
  effects = list(iset,
                 data.frame(intercept = 1,
                            x1 = inla_df$x1,
                            x2 = inla_df$x2)),
  tag     = "est"
)

coords_pred <- as.matrix(env_valid[, c("x", "y")])

# Projector matrix for prediction locations
A_pred <- inla.spde.make.A(mesh, loc = coords_pred)

stack_pred <- inla.stack(
  data    = list(y = NA),
  A       = list(A_pred, 1),
  effects = list(
    iset,
    data.frame(intercept = 1,
               x1 = env_valid$temp,
               x2 = env_valid$prec)
  ),
  tag = "pred"
)

# Join estimation + prediction stacks
stack_full <- inla.stack(stack_est, stack_pred)

```



```{r 999pred-INLA}

# Refit on full stack
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),

  control.fixed = list(
    mean = list(intercept = prior_alpha_mean,
                x1        = prior_beta1_mean,
                x2        = prior_beta2_mean),
    prec = list(intercept = prec_alpha,
                x1        = prec_beta1,
                x2        = prec_beta2)
  ),

  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]
```


```{r 999Plot-INLA}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Model update - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)

png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Model update - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()
```


# INLA model - non informed priors

```{r 999prior-NonInfo}

prior.beta <- list(prec = list(default = 1/(1)^2, Intercept = 1/(20)^2), 
                   mean = list(default = 0, Intercept = 0))

```



```{r 999full-NonInfo}
fit_inla_full <- inla(
  formula,
  family = "binomial",
  data   = inla.stack.data(stack_full),
  control.fixed = prior.beta,
  control.predictor = list(
    A       = inla.stack.A(stack_full),
    compute = TRUE,
    link    = 1
  ),

  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE)
)

# Extract predictions for pred stack only
idx_pred   <- inla.stack.index(stack_full, tag = "pred")$data

pred_mean  <- fit_inla_full$summary.fitted.values[idx_pred, "mean"]
pred_sd    <- fit_inla_full$summary.fitted.values[idx_pred, "sd"]


```


```{r 999Plot--NonInfo}
pred_raster <- rast(env$temp)
values(pred_raster) <- NA
pred_raster[valid_idx] <- pred_mean

sd_raster <- rast(env$temp)
values(sd_raster) <- NA
sd_raster[valid_idx] <- pred_sd

non_na_extent <- ext(trim(pred_raster))
pred_cropped <- crop(pred_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_pred_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(pred_cropped,
     main = "Non informative - Species Distribution Model",
     col  = hcl.colors(100, "YlGnBu"),
     range = c(0, 1),
     axes   = FALSE,
     box    = FALSE)
dev.off()

non_na_extent <- ext(trim(sd_raster))
sd_cropped <- crop(sd_raster, non_na_extent)
png(paste0("C:/Users/maria/Documents/Chapter1/bayesian_model/results/", "NI_SD_", removal_pct_txt, ".png"), width = 800, height = 800, res = 120)
plot(sd_cropped,
     main = "Non informative - Uncertainty (posterior SD)",
     col  = hcl.colors(100, "Reds"),
     axes   = FALSE,
     box    = FALSE)
dev.off()

```

```{r }

```